---
title: "Creating metacommunities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating metacommunities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(flume)
devtools::load_all()  # for dev, remove when stable version done
```


The metacommunity is one of the two main components for creating a `flume` model. It describes the 
collection of species that can occur in the model, including all of their traits related to the 
niche, species interactions, and dispersal.

For all pieces of information that are required, there is also default behaviour. In this way, a 
metacommunity can be generated with real species in mind, or completely randomly, or something in 
between. It is useful to check out the help file to understand the defaults for the various options.

## Niche shape

The niche contains two components, a colonisation and an extinction function. Currently, the only 
implemented shape is a `gaussian` colonisation function and a `constant` extinction function. Each 
of them has parameters:

### Gaussian
  * scale: the maximum value of the curve
  * location: where on the resource axis the peak of the curve is
  * width: How broad (i.e., the standard deviation of the gaussian function)
  
### Constant
  * scale: the height of the function

These parameters can be specified to vary by species, vary by resource axis, or both, and to have 
orthogonal or covarying resource axes. The help file explains how to provide each of these cases.


## Dispersal parameters

Species can disperse in the river network in two ways. **Active dispersal**, controlled by a
parameter `alpha`, indicates the ability of a species to move under its own power; active dispersal
occurs along the stream, in both up- and down-stream directions. Passive dispersal that occurs
when one species is transported by another species should also be included in `alpha`. **Passive
dispersal** is dispersal with the flow of water, and only occurs downstream. It is specified with a
parameter `beta`. Both `alpha` and  `beta` can be a scalar (all species have the same disperal
ability) or a vector with one entry per species.


## Resource use

A single parameter, `r_scale`, controls the resource use of species. Generally speaking, species 
consume resources faster closer to their niche optimum, scaled by this parameter. Like the niche
parameters, resource use can vary by species, resource, or both.


## Generating metacommunities

### Creating metacommunities by hand

Here we generate a metacommunity for algae by hand. *nb use the stuff from the thomas_flume_ex.r*

```{r algae}
# niche locations
niche_loc = c(ASV_4 = 200.51879, ASV_5 = 200.51879, ASV_17 = 139.15023, ASV_63 = 18.47656,
	ASV_67 = 414.07093)

# niche widths
niche_wid = c(ASV_4 = 164.629301, ASV_5 = 164.629301, ASV_17 = 102.373506, ASV_63 = 3.765155,
	ASV_67 = 160.079609)

mc = metacommunity(niche_loc, niche_wid, niche_lim = c(0, 650))
```


### Creating metacommunities using scenarios