---
title: "Creating metacommunities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating metacommunities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(flume)
# devtools::load_all()  # for dev, remove when stable version done
```

The metacommunity is one of the two main components for creating a `flume` model. It describes the 
collection of species that can occur in the model, including all of their traits related to the 
niche, species interactions, and dispersal.

The easiest way to generate these components is by using `scenarios`, which you pass to the
function `metacommunity()`. There are scenarios around resources and niches, and scenarios for
dispersal. For example, the default behaviour when you call the `metacommunity()` function with no
arguments is to create a metacommunity with 2 species and one resource using the `niches_uniform()`
scenario for resources/niches and the `dispersal_custom()` dispersal scenario (using default
parameters in this case that do not vary by species). The equivalent call specifying all parameters
would be: `metacommunity(nsp = 2, nr = 1, niches = niches_uniform, dispersal = dispersal_custom)`.
The niche scenario here will create niches that are uniformly spaced across the possible resource
values, with a standard deviation chosen automatically to cover a moderate breadth.

The focus of this vignette will be explaining the options that can be invoked, primarily using the
custom scenarios. In these scenarios, or all pieces of information that can be provided required,
there are also defaultsr. In this way, a  metacommunity can be generated with real species in mind,
or completely randomly, or something in between. It is useful to check out the help file to
understand the defaults for the various options.

## Niche shape

The niche contains two components, a colonisation and an extinction function. Currently, the only 
implemented shape is a `gaussian` colonisation function and a `constant` extinction function. Each 
of them has parameters:

### Gaussian
  * scale: the maximum value of the curve
  * location: where on the resource axis the peak of the curve is
  * breadth: How broad (i.e., the standard deviation of the gaussian function)
  
### Constant
  * scale: the height of the function

These parameters can be specified to vary by species, vary by resource axis, or both, and to have 
orthogonal or covarying resource axes. To change them individually, use the `niches_custom`
scenario. Here, we change the location and breadth of the gaussian colonisation function to make
three species, including two low-resource specialists that compete strongly and one generalist with
a slight preference for high resource conditions, but a lower peak performance.

```{r niche, fig.width = 9, fig.height = 3.5}
nopts = list(location = c(0.1, 0.12, 0.7), breadth = c(0.1, 0.1, 2), scale_c = c(0.5, 0.5, 0.25))
mcom = metacommunity(nsp = 3, nr = 1, niches = niches_custom, niche_args = nopts)
plot(mcom)
```


## Dispersal parameters

Species can disperse in the river network in two ways. **Active dispersal**, controlled by a
parameter `alpha`, indicates the ability of a species to move under its own power; active dispersal
occurs along the stream, in both up- and down-stream directions. Passive dispersal that occurs
when one species is transported by another species should also be included in `alpha`. **Passive
dispersal** is dispersal with the flow of water, and only occurs downstream. It is specified with a
parameter `beta`. Both `alpha` and  `beta` can be a scalar (all species have the same disperal
ability) or a vector with one entry per species. We can change these parameters using the
`dispersal_custom` scenario, in a similar manner as to the niches. Here we edit the previous
community to make the specialists primarily passive dispersers, while the generalist is a strong
disperser both up- and downstream.

```{r dispersal, fig.width = 9, fig.height = 3.5}
dopts = list(alpha = c(0.05, 0.05, 0.5), beta = 0.5)
mcom = metacommunity(nsp = 3, nr = 1, niches = niches_custom, dispersal = dispersal_custom,
  niche_args = nopts, dispersal_args = dopts)
```


### Complex resources

It isn't always the case that resources map directly to niche axes. Here we load the `algae` dataset
included with `flume`. The `niches` object gives average and standard deviation of N:P ratios for 5
aquatic algal taxa. We can create a metacommunity that tracks two resources, but computes the niche
based on the ratio of these two resources.

```{r algae, fig.width = 9, fig.height = 3.5}
data(algae)

nopts = list(location = algae$niches$location, breadth = algae$niches$breadth, ratio = c(1, 2))
mcom = metacommunity(nsp = nrow(algae$niches), nr = 2, niches = niches_custom, niche_args = nopts,
  sp_names = algae$niches$species, r_names = c("N", "P"))
plot(mc, res = 2000)
```

