---
title: "A simple simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A simple simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 5.5, fig.height = 5.5, collapse = TRUE, comment = "##")

library(flume)
```



Here we demonstrate the model using a simple simulation with generated data. There are several 
essential components:

1. A `metacommunity` object, which describes all the possible species that can occur in the 
simulation, gives their niches, and includes the pairwise competition strength.
2. A `river_network`, which describes the topology of the network and gives other important 
attributes, such as discharge.
3. State variables; these are (mostly) dynamic attributes describing the state of the model at a 
given time. 
4. Boundary conditions: input, both to the biological community 
(via immigration from the surrounding landscape) and to the resources (via terrestrial runoff, 
groundwater intrusion, etc).


## 1. The metacommunity

The metacommunity is an object created by `flume` that describes all of the species that can
possibly occur in the model. It also describes the possible resources and other habitat variables,
their range of possible values, how the resources behave, the species' niches, how species
compete for resources, and species' dispersal abilities.

Here we generate a metacommunity using a scenario. It is also possible to have finer control over
how the metacommunity is modelled; see the "[metacommunities](metacommunities.html)" vignette for 
more details.


```{r fig.height = 3, fig.width = 7}
mc = metacommunity(nsp = 3, nr = 1, niches = niches_uniform)
plot(mc)
```

The `niches_uniform` scenario spreads the resources out along a gradient, with species' optima 
uniformly spread along this gradient. We can see the niche parameters used with the `niche_par`
function:

```{r}
cbind(
  niche_par(mc, "location"),
  niche_par(mc, "sd")
)
```




## 2. The river network

The main component of the river network is a weighted digraph; i.e., the topological relationships
between locations. The network graph in `flume` uses equal-length stream reaches as nodes, while
the edges describe which reaches are neighbours of each other. The simplest way to define a network
is using an attribute list and a weighted adjacency matrix.

Discharge and cross-sectional area are required attributes (although area can be estimated from Q
if omitted). Here, Q grows by 1 each time we move downstream (i.e., when we traverse an edge,
representing lateral input), and we add reaches together at confluences. We define node 1 to be the
most downstream node (this is purely an arbitrary choice). We define a river network with two
equal subcatchments and a mainstem.

```{r}
Q1 = Q2 = c(13,12,8,7,1,4,1,1,2,1)
Q3 = c(31,30,29,28,27)
```

The adjacency matrix describes the spatial relationships among nodes. The matrix must be square,
with one row and column per node. For two sites `i` and `j`, `adj[i,j]` will be nonzero if `i` is
upstream of `j`. The layout matrix shows where to plot each vertex; this is optional but can be
useful for nicer plots.

It should be clear that creating a non-trivial network by hand is cumbersome! Fortunately there is
a way to do this for a real river using a digital elevation model; see the `network_import` vignette
for more details.



```{r}
## matrix for the two subcatchments
nsites = length(Q1)
adj = matrix(0, nrow=nsites, ncol=nsites)
adj[2,1] = adj[3,2] = adj[9,2] = adj[4,3] = adj[5,4] = adj[6,4] = adj[7,6] = adj[8,6] = 
  adj[10,9] = 1
adj2 = adj

# layout is optional, and controls where to place the nodes
# here we just use values to make the network look somewhat nice
# For a real river, x-y spatial coordinates for each reach chould be used
layout1 = matrix(c(0,0,0,1,-0.5,2,-0.5,3,-1,4,0,4,-0.5,5,0.5,5,0.5,2,1,3), byrow=TRUE, nrow=nsites)
layout2 = layout1
layout2[,1] = layout2[,1] + 3

rownames(adj) = colnames(adj) = paste("c1", 1:nrow(adj), sep="_")
rownames(adj2) = colnames(adj2) = paste("c2", 1:nrow(adj2), sep="_")
nsites3 = length(Q3)
adj3 = matrix(0, nrow=nsites3, ncol=nsites3)
adj3[2,1] = adj3[3,2] = adj3[4,3] = adj3[5,4] = 1
rownames(adj3) = colnames(adj3) = paste("c3", 1:nrow(adj3), sep="_")
layout3 = matrix(c(rep(1.5, nsites3), (-nsites3):(-1)), ncol=2)

# now create an adjacency matrix for all subcatchments together
adj_all = matrix(0, nrow = nsites+nsites+nsites3, ncol = nsites+nsites+nsites3)
rownames(adj_all) = colnames(adj_all) = c(rownames(adj), rownames(adj2), rownames(adj3))

# add the links from the other matrices
adj_all[rownames(adj), colnames(adj)] = adj
adj_all[rownames(adj2), colnames(adj2)] = adj2
adj_all[rownames(adj3), colnames(adj3)] = adj3

# link catchments 1 and 2 to catchment 3
adj_all['c1_1', 'c3_5'] = adj_all['c2_1', 'c3_5'] = 1

network = river_network(adjacency = adj_all, discharge = c(Q1, Q2, Q3), 
  layout = rbind(layout1, layout2, layout3))
```

By default, plotting the network will show the edges weighted by discharge. Node ('vertex') colours
can be set by `vertex.color`; I use a simple gradient following the order of sites, because this is
also how I will set up my resource gradients (in **3. Setting up the model** below). Edge weight is
proportional to the discharge between sites. 

```{r fig.height = 7, fig.width = 7}
plot(network, edge.arrow.size = 0.5, edge.width = 20, vertex.label.cex = 0.7)
```


## 3. Starting conditions

We will now set the starting resource conditions and species distributions. For this example,
we have 3 species along a resource gradient (where the resource values are between 0 and 1). We set
up the network here so that there is a spatial gradient in resources, with high concentration in one
subcatchment, low in the other, and intermediate values in the mainstem.  We then use an equilibrium 
scenario to establish initial species distributions.

```{r fig.height = 7, fig.width = 7}
st0 = rep(0.4, nrow(adj_all))

## headwaters at opposite sides of the gradient, then gradually come together moving downstream
st0[c(5,7,8,10)] = 0.85
st0[c(5,7,8,10)+10] = 0.1
st0[c(3,4,6,9)] = 0.75
st0[c(3,4,6,9)+10] = 0.15
st0[c(1:2)] = 0.7
st0[c(1:2)+10] = 0.2
st0 = matrix(st0, ncol=1)

state(network) = st0
plot(network, edge.arrow.size = 0.5, edge.width = 20, vertex.label.cex = 0.7)
```


```{r fig.height = 5, fig.width = 15}
sp0 = community_equilibrium(network, mc)
site_by_species(network) = sp0
plot(network, "site_by_species", edge.arrow.size = 0, edge.width = 20, vertex.label.cex = 0.7)
```




## 3. Setting up the model

The model is initialised with the `flume()` function. Here, we should also specify initial values
for the two state variables (site by resource concentration, and site by species).

The state variables are a matrix, one row per node in the river network, with each column
representing either resources or species. The resource state variable can be accessed with
`state(network)`, and the species with `site_by_species(network)`.

By default, the boundary condition for resources will use identical values as the initial state, and
the boundary condition for species is set to zero. These can be changed using the `stb` and `spb`
arguments.

We use a scenario for the initial site by species matrix that places species at any site where they
would be expected to be found at equilibrium in the absence of competition. This requires the
initial state of the network to be specified in advance. Scenarios for resources
are coming in a future version; for now we simply set an upstream to downstream gradient.

```{r}
model = flume(mc, network, sp0, st0)
```


## 4. Running the model

Now we are prepared to run the simulation. Here we run for 100 time steps and then explore some of
the output options. In this case, all species survive the entire simulation.

```{r fig.height = 7, fig.width = 7}
set.seed(123)
model_run = run_simulation(model, 100)

## occupancy plot
plot(model_run)
```

We can see that resources find an equilibrium after a few time steps. Resource consumption is quite
low in this scenario, so resource concentrations are determined by transport and lateral input.

```{r fig.height = 7, fig.width = 7}
## resource plot
plot(model_run, variable = "resources")
```

Here we can see the species distributions at the end of the simulation. Note that flume objects
store the network in a list (to allow for duplicate networks to be run under the same conditions),
hence the `[[1]]` index to plot the network.

```{r fig.height = 5, fig.width = 15}
## species distributions at the end
plot(model_run$network[[1]], variable = "site_by_species")
```

```{r}
## additional plots -- forthcoming

## occupancy in time by reach
## this should be a map of the network, colored so that
## each node shows the proportion of time it was occupied by each species

## diversity over time
## diversity in space
## EF over time
## EF over space
## EF vs transport


```

